Tropical Andes Taxonomic Diversity of Plants and Frugivores

### 2024-06-17 code/L2/L2_1_TropicalAndes_taxdiv.Rmd
add code for parallel processing, save individual plots to file

---
title: "Tropical Andes Taxonomic Diversity of Plants and Frugivores"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "Calculating and mapping taxonomic diversity as species richness using occurrence data for plants and frugivores in the Tropical Andes Moist Lowland and Montane forests. Note: some code is adapted from https://luisdva.github.io/rstats/richness/."
data input: "TropicalAndes_plant_occ.csv, TropicalAndes_frugivore_occ.csv"
data output: "TropicalAndes_frugivore_richness.csv, TropicalAndes_GBIF_plant_richness.csv"
date: "2023-08-01"
output: html_document
---

# Load required packages
```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(ggspatial)
library(ggpubr)
```

# Set file paths
```{r}
data_path_L0<-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L0')
data_path_L1 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L2')
figure_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/figures')
```

@@

# Plot TD/Richness
## Function for richness plots
```{r}
create_rich_plots <- function(resolution_meters) {
 # Make Grid
  TAGrid <- TApoly %>%
    st_make_grid(cellsize = c(resolution_meters)) %>%
    st_intersection(TropicalAndes_IUCNHabitat_Forest) %>%
    st_cast("MULTIPOLYGON") %>%
    st_sf() %>%
    mutate(cellid = row_number())

  # Function to calculate richness for a given dataset
  calculate_richness <- function(data_sf, species_sf) {
    result <- data_sf %>%
      st_join(species_sf) %>%
      mutate(overlap = ifelse(!is.na(species), 1, 0)) %>%
      group_by(cellid) %>%
      summarize(num_species = sum(overlap)) %>%
      ungroup()
    return(result)
  }

@@

  # Split the TAGrid into chunks if necessary
  TAGrid_chunks <- split(TAGrid, ceiling(seq_along(TAGrid$cellid) / chunk_size))

  # Initialize empty data frames for results
  plant_richness_grid <- list()
  frugivore_richness_grid <- list()

  # Iterate over chunks to calculate richness
  for (i in seq_along(TAGrid_chunks)) {
    chunk <- TAGrid_chunks[[i]]

    # Debug: Print chunk info
    print(paste("Processing chunk", i, "of", length(TAGrid_chunks)))

    plant_richness_chunk <- calculate_richness(chunk, plants_sf_species)
    frugivore_richness_chunk <- calculate_richness(chunk, frugivores_sf_species)

    # Store the chunks in lists
    plant_richness_grid[[i]] <- plant_richness_chunk
    frugivore_richness_grid[[i]] <- frugivore_richness_chunk
  }

  # Combine all chunks into final data frames
  plant_richness_grid <- bind_rows(plant_richness_grid)
  frugivore_richness_grid <- bind_rows(frugivore_richness_grid)

  # Helper functions to generate plots
  generate_plant_plot <- function(data) {
    ggplot(data) +
      geom_sf(data = Americas, fill = "white") +
      geom_sf(data = TApoly, fill = "grey", size = 0.1) +
      geom_sf(aes(fill = num_species), color = NA) +
      scale_fill_distiller(palette = "Greens", direction = 1, limits = c(0,6000)) +
      labs( x = "Longitude", y = "Latitude", fill = "Number of species") +
      coord_sf(xlim = c(-82, -60), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
      annotation_scale(location = "bl", width_hint = 0.5)  +
      theme(panel.background = element_rect(fill = "lightblue"))
  }
  generate_frugivore_plot <- function(data) {
    ggplot(data) +
      geom_sf(data = Americas, fill = "white") +
      geom_sf(data = TApoly, fill = "grey", size = 0.1) +
      geom_sf(aes(fill = num_species), color = NA) +
      scale_fill_distiller(palette = "YlOrBr", direction = 1, limits = c(0,500)) + 
      labs(x = "Longitude", y = "Latitude", fill = "Number of species") +
      coord_sf(xlim = c(-82, -60), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
      annotation_scale(location = "bl", width_hint = 0.5)  +
      theme(panel.background = element_rect(fill = "lightblue"))
  }

@@

  # Create histograms
  plant_richness_hist <- hist(plant_richness_grid$num_species, 
    main = paste("Histogram of Plant Richness [", resolution_meters/1000, " km] by cell", sep = ""),
                              xlab = "Plant Richness by cell")
  frugivore_richness_hist <- hist(frugivore_richness_grid$num_species,
    main = paste("Histogram of Frugivore Richness [", resolution_meters/1000, " km] by cell", sep = ""),
                                  xlab = "Frugivore Richness by cell")
  # Return a list of data
  list(plantgridRichTA = plant_plot,
       frugivoregridRichTA = frugivore_plot,
       plant_cellRichness = plant_richness_grid,
       frugivore_cellRichness = frugivore_richness_grid, 
       plant_richness_hist = plant_richness_hist,
       frugivore_richness_hist = frugivore_richness_hist)
}

```

## 5 km resolution
```{r}
richness_5km <- create_rich_plots(5000)
print(richness_5km)

@@

## 10 km
```{r}
richness_10km <- create_rich_plots(10000)
print(richness_10km)

@@

## 25 km
```{r}
richness_25km <- create_rich_plots(25000)
print(richness_25km)

@@

## 50 km
```{r}
richness_50km <- create_rich_plots(50000)
print(richness_50km)
# save plots
plantgridRichTA_50km <- richness_50km$plantgridRichTA
frugivoregridRichTA_50km <- richness_50km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_50km <- richness_50km$plant_cellRichness
frugivore_cellRichness_50km <- richness_50km$frugivore_cellRichness
```

## 75 km 
```{r}
richness_75km <- create_rich_plots(75000)
print(richness_75km)
# save plots
plantgridRichTA_75km <- richness_75km$plantgridRichTA
frugivoregridRichTA_75km <- richness_75km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_75km <- richness_75km$plant_cellRichness
frugivore_cellRichness_75km <- richness_75km$frugivore_cellRichness
```

## 100 km 
```{r}
richness_100km <- create_rich_plots(100000)
print(richness_100km)
# save plots
plantgridRichTA_100km <- richness_100km$plantgridRichTA
frugivoregridRichTA_100km <- richness_100km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_100km <- richness_100km$plant_cellRichness
frugivore_cellRichness_100km <- richness_100km$frugivore_cellRichness
```

# Multi panel plots
## Richness maps
```{r}
all_plant_richness_plots <- ggpubr::ggarrange(# plantgridRichTA_5km, 
                                              plantgridRichTA_10km, plantgridRichTA_25km,
                                              plantgridRichTA_50km, plantgridRichTA_75km,
                                              plantgridRichTA_100km,
                                              ncol = 5, nrow = 1,
                                              labels = c("10 km", "25 km", "50 km", "75 km", "100 km"),
                                              common.legend = TRUE, legend = "right")
all_plant_richness_plots
all_plant_richness_plots_labeled <- ggpubr::annotate_figure(all_plant_richness_plots,
                                                            left = ggpubr::text_grob("Plants", face = "bold", size = 14))
all_plant_richness_plots_labeled
```

```{r}
all_frugivore_richness_plots <- ggpubr::ggarrange(#frugivoregridRichTA_5km, 
                                              frugivoregridRichTA_10km, frugivoregridRichTA_25km,
                                              frugivoregridRichTA_50km, frugivoregridRichTA_75km,
                                              frugivoregridRichTA_100km,
                                              ncol = 5, nrow = 1,
                                              labels = c("10 km", "25 km", "50 km", "75 km", "100 km"),
                                              common.legend = TRUE, legend = "right")
all_frugivore_richness_plots_labeled <- ggpubr::annotate_figure(all_frugivore_richness_plots,
                                                            left = ggpubr::text_grob("Frugivores", face = "bold", size = 14))
all_frugivore_richness_plots_labeled
```

```{r}
all_richness_plots <- ggpubr::ggarrange(all_plant_richness_plots_labeled,
                                        all_frugivore_richness_plots_labeled,
                                        ncol = 1, nrow = 2)
all_richness_plots
ggsave("all_richness_plots.png", all_richness_plots, path = figure_path, width = 11, height = 8, units = "in")
```

# Write data to csv

```{r}
write.csv(plant_cellRichness_5km, file.path(output_path,"TropicalAndes_plantRichness_5km.csv"))
write.csv(frugivore_cellRichness_5km, file.path(output_path,"TropicalAndes_frugivoreRichness_5km.csv"))
```

>>>>

---
title: "Tropical Andes Taxonomic Diversity of Plants and Frugivores"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "Calculating and mapping taxonomic diversity as species richness using occurrence data for plants and frugivores in the Tropical Andes Moist Lowland and Montane forests. Note: some code is adapted from https://luisdva.github.io/rstats/richness/."
#data input: "TropicalAndes_plant_occ.csv, TropicalAndes_frugivore_occ.csv"
#data output: "TropicalAndes_frugivore_richness.csv, TropicalAndes_GBIF_plant_richness.csv"
date: "2023-08-01"
output: 
  html_document:
    highlight_downlit = TRUE
    df_print: tibble
---

# Load required packages
```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(parallel)
library(foreach)
library(doParallel)
library(ggspatial)
library(ggpubr)
```

# Set file paths
```{r}
data_path_L0 <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L0')
data_path_L1 <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L2')
figure_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/figures')
```

@@

# Plot TD/Richness
## Function for richness plots using parallel processing
```{r}
# Define the calculate_richness function outside
calculate_richness <- function(data_sf, species_sf) {
  result <- data_sf %>%
    st_join(species_sf) %>%
    mutate(overlap = ifelse(!is.na(species), 1, 0)) %>%
    group_by(cellid) %>%
    summarize(num_species = sum(overlap)) %>%
    ungroup()
  return(result)
}

create_rich_plots <- function(resolution_meters) {
  # Make Grid
  TAGrid <- TApoly %>%
    st_make_grid(cellsize = c(resolution_meters)) %>%
    st_intersection(TropicalAndes_IUCNHabitat_Forest) %>%
    st_cast("MULTIPOLYGON") %>%
    st_sf() %>%
    mutate(cellid = row_number())

@@

  # Split the TAGrid into chunks
  TAGrid_chunks <- split(TAGrid, ceiling(seq_along(TAGrid$cellid) / chunk_size))

  # Set up parallel processing
  no_cores <- detectCores() - 1
  cl <- makeCluster(no_cores)
  registerDoParallel(cl)

  # Export necessary objects to the cluster
  clusterExport(cl, varlist = c("plants_sf_species", "frugivores_sf_species", "calculate_richness"))

  # Process chunks in parallel
  plant_richness_grid <- foreach(chunk = TAGrid_chunks, .combine = bind_rows, .packages = c("dplyr", "sf")) %dopar% {
    calculate_richness(chunk, plants_sf_species)
  }

  frugivore_richness_grid <- foreach(chunk = TAGrid_chunks, .combine = bind_rows, .packages = c("dplyr", "sf")) %dopar% {
    calculate_richness(chunk, frugivores_sf_species)
  }

  # Stop the parallel cluster
  stopCluster(cl)

  # Helper functions to generate plots
  generate_plant_plot <- function(data) {
    ggplot(data) +
      geom_sf(data = Americas, fill = "white") +
      geom_sf(data = TApoly, fill = "grey", size = 0.1) +
      geom_sf(aes(fill = num_species), color = NA) +
      scale_fill_distiller(palette = "Greens", direction = 1, limits = c(0,1200)) +
      labs(x = "Longitude", y = "Latitude", fill = "S") +
      coord_sf(xlim = c(-82, -60), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
      annotation_scale(location = "bl", width_hint = 0.5) +
      theme(panel.background = element_rect(fill = "lightblue"))
  }

  generate_frugivore_plot <- function(data) {
    ggplot(data) +
      geom_sf(data = Americas, fill = "white") +
      geom_sf(data = TApoly, fill = "grey", size = 0.1) +
      geom_sf(aes(fill = num_species), color = NA) +
      scale_fill_distiller(palette = "YlOrBr", direction = 1, limits = c(0,500)) +
      labs(x = "Longitude", y = "Latitude", fill = "S") +
      coord_sf(xlim = c(-82, -60), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
      annotation_scale(location = "bl", width_hint = 0.5) +
      theme(panel.background = element_rect(fill = "lightblue"))
  }

@@

  # Create histograms
  plant_richness_hist <- hist(plant_richness_grid$num_species,
                              main = paste("Histogram of Plant Richness [", resolution_meters / 1000, " km] by cell", sep = ""),
                              xlab = "Plant Richness by cell")
  frugivore_richness_hist <- hist(frugivore_richness_grid$num_species,
                                  main = paste("Histogram of Frugivore Richness [", resolution_meters / 1000, " km] by cell", sep = ""),
                                  xlab = "Frugivore Richness by cell")

  # Return a list of data
  list(plantgridRichTA = plant_plot,
       frugivoregridRichTA = frugivore_plot,
       plant_cellRichness = plant_richness_grid,
       frugivore_cellRichness = frugivore_richness_grid,
       plant_richness_hist = plant_richness_hist,
       frugivore_richness_hist = frugivore_richness_hist)
}
```

## 5 km resolution
```{r}
richness_5km <- create_rich_plots(5000)
```

```{r}
print(richness_5km)

@@

## 10 km
```{r}
richness_10km <- create_rich_plots(10000)
```

```{r}
print(richness_10km)

@@

## 25 km
```{r}
richness_25km <- create_rich_plots(25000)
```

```{r}
print(richness_25km)

@@

## 50 km
```{r}
richness_50km <- create_rich_plots(50000)
```

```{r}
print(richness_50km)
# save plots
plantgridRichTA_50km <- richness_50km$plantgridRichTA
frugivoregridRichTA_50km <- richness_50km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_50km <- richness_50km$plant_cellRichness
frugivore_cellRichness_50km <- richness_50km$frugivore_cellRichness
```

## 75 km 
```{r}
richness_75km <- create_rich_plots(75000)
```

```{r}
print(richness_75km)
# save plots
plantgridRichTA_75km <- richness_75km$plantgridRichTA
frugivoregridRichTA_75km <- richness_75km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_75km <- richness_75km$plant_cellRichness
frugivore_cellRichness_75km <- richness_75km$frugivore_cellRichness
```

## 100 km 
```{r}
richness_100km <- create_rich_plots(100000)
```

```{r}
print(richness_100km)
# save plots
plantgridRichTA_100km <- richness_100km$plantgridRichTA
frugivoregridRichTA_100km <- richness_100km$frugivoregridRichTA
## Extract cell values
plant_cellRichness_100km <- richness_100km$plant_cellRichness
frugivore_cellRichness_100km <- richness_100km$frugivore_cellRichness
```

# Save richness objects to file
```{r}
saveRDS(richness_5km, file = file.path(output_path,"richness_5km.rds"))
saveRDS(richness_10km, file = file.path(output_path,"richness_10km.rds"))
saveRDS(richness_25km, file = file.path(output_path,"richness_25km.rds"))
saveRDS(richness_50km, file = file.path(output_path,"richness_50km.rds"))
saveRDS(richness_75km, file = file.path(output_path,"richness_75km.rds"))
saveRDS(richness_100km, file = file.path(output_path,"richness_100km.rds"))
```

# Save plots to file
```{r}
plot_list <- list(plantgridRichTA_5km = plantgridRichTA_5km, plantgridRichTA_10km = plantgridRichTA_10km,
                  plantgridRichTA_25km = plantgridRichTA_25km, plantgridRichTA_50km = plantgridRichTA_50km,
                  plantgridRichTA_75km = plantgridRichTA_75km, plantgridRichTA_100km = plantgridRichTA_100km,
                  frugivoregridRichTA_5km = frugivoregridRichTA_5km, frugivoregridRichTA_10km = frugivoregridRichTA_10km,
                  frugivoregridRichTA_25km = frugivoregridRichTA_25km, frugivoregridRichTA_50km = frugivoregridRichTA_50km,
                  frugivoregridRichTA_75km = frugivoregridRichTA_75km, frugivoregridRichTA_100km = frugivoregridRichTA_100km)
#histogram_list <- list(plant_richness_hist_5km = richness_5km$plant_richness_hist, 
#                       frugivore_richness_hist_5km = richness_5km$frugivore_richness_hist,
#                       plant_richness_hist_10km = richness_10km$plant_richness_hist,
#                       frugivore_richness_hist_10km = richness_10km$frugivore_richness_hist,
#                       plant_richness_hist_25km = richness_25km$plant_richness_hist,
#                       frugivore_richness_hist_55km = richness_25km$frugivore_richness_hist,
#                       plant_richness_hist_50km = richness_50km$plant_richness_hist,
#                       frugivore_richness_hist_50km = richness_50km$frugivore_richness_hist,
#                       plant_richness_hist_75km = richness_75km$plant_richness_hist,
#                       frugivore_richness_hist_75km = richness_75km$frugivore_richness_hist,
#                       plant_richness_hist_100km = richness_100km$plant_richness_hist,
#                       frugivore_richness_hist_100km = richness_100km$frugivore_richness_hist)
#histograms would have to be converted to ggplot to save
```

```{r}
save_plots <- function(plot_list, figure_path) {
  # Iterate over the list and save each plot
  for (plot_name in names(plot_list)) {
    plot_path <- file.path(figure_path, paste0(plot_name, ".png"))

    # Try to save the plot and handle any potential errors
    tryCatch({
      ggsave(filename = plot_path, 
             plot = plot_list[[plot_name]])
      cat("Saved plot:", plot_path, "\n")
    }, error = function(e) {
      cat("Failed to save plot:", plot_name, "\nError message:", e$message, "\n")
    })
  }
}
```

```{r}
# Save the plots
save_plots(plot_list, figure_path)
```

# Multi panel plots of richness maps
## Plants
```{r}
# Add labels and remove axis labels from all plots
plantgridRichTA_5km_labeled <- plantgridRichTA_5km + ggtitle("5 km") + theme(plot.title = element_text(hjust = 0.5)) + theme(axis.title = element_blank())
plantgridRichTA_10km_labeled <- plantgridRichTA_10km + ggtitle("10 km") + theme(plot.title = element_text(hjust = 0.5)) + theme(axis.title = element_blank())
plantgridRichTA_25km_labeled <- plantgridRichTA_25km + ggtitle("25 km") + theme(plot.title = element_text(hjust = 0.5)) + theme(axis.title = element_blank())
plantgridRichTA_50km_labeled <- plantgridRichTA_50km + ggtitle("50 km") + theme(plot.title = element_text(hjust = 0.5)) + theme(axis.title = element_blank())
plantgridRichTA_75km_labeled <- plantgridRichTA_75km + ggtitle("75 km") + theme(plot.title = element_text(hjust = 0.5)) + theme(axis.title = element_blank())
plantgridRichTA_100km_labeled <- plantgridRichTA_100km + ggtitle("100 km") + theme(plot.title = element_text(hjust = 0.5)) + theme(axis.title = element_blank())
plantgridRichTA_100km_labeled
```

```{r}
# Arrange the plots with common legend
all_plant_richness_plots <- ggarrange(plantgridRichTA_5km_labeled, plantgridRichTA_10km_labeled,
                                      plantgridRichTA_25km_labeled, plantgridRichTA_50km_labeled,
                                      plantgridRichTA_75km_labeled, plantgridRichTA_100km_labeled,
                                      ncol = 6, nrow = 1,
                                      common.legend = TRUE, legend = "left")

all_plant_richness_plots
```

```{r}
# Add "Plants" label above the legend
all_plant_richness_plots_labeled <- all_plant_richness_plots + theme(legend.title = element_text("Plants\n S"))
all_plant_richness_plots_labeled
ggsave("plant_richness_plots.png", all_plant_richness_plots_labeled, path = figure_path, width = 18, height = 8, units = "in")
```

## Frugivores
```{r}
all_frugivore_richness_plots <- ggpubr::ggarrange(#frugivoregridRichTA_5km, 
                                              frugivoregridRichTA_10km, frugivoregridRichTA_25km,
                                              frugivoregridRichTA_50km, frugivoregridRichTA_75km,
                                              frugivoregridRichTA_100km,
                                              ncol = 5, nrow = 1,
                                              labels = c("10 km", "25 km", "50 km", "75 km", "100 km"),
                                              common.legend = TRUE, legend = "right")
```

```{r}
all_frugivore_richness_plots_labeled <- ggpubr::annotate_figure(all_frugivore_richness_plots,
                                                            left = ggpubr::text_grob("Frugivores", face = "bold", size = 14))
all_frugivore_richness_plots_labeled
ggsave("frugivore_richness_plots.png", all_frugivore_richness_plots_labeled, path = figure_path, width = 11, height = 8, units = "in")
```

## Combine plants & frugivores
```{r}
all_richness_plots <- ggpubr::ggarrange(all_plant_richness_plots_labeled,
                                        all_frugivore_richness_plots_labeled,
                                        ncol = 1, nrow = 2)
all_richness_plots
ggsave("all_richness_plots.png", all_richness_plots, path = figure_path, width = 11, height = 8, units = "in")
```

# Write data to csv
```{r}
write.csv(plant_cellRichness_5km, file.path(output_path,"TropicalAndes_plantRichness_5km.csv"))
write.csv(frugivore_cellRichness_5km, file.path(output_path,"TropicalAndes_frugivoreRichness_5km.csv"))
```
